---
title: "Brain Extraction and Segmentation"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brain Extraction and Segmentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
library(kirby21.t1)
library(fslr)
library(neurobase)
library(extrantsr)
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/neuroc/master/brain_extraction/index.R).

In this tutorial we will discuss performing brain segmentation using the brain extraction tool (BET) in `fsl` and a robust version using a wrapper function in `extrantsr`, `fslbet_robust`. 

# Data Packages

For this analysis, I will use one subject from the Kirby 21 data set.  The `kirby21.base` and `kirby21.t1` packages are necessary for this analysis and have the data we will be working on.  You need devtools to install these.  Please refer to [installing devtools](../installing_devtools/index.html) for additional instructions or troubleshooting.


```{r, eval = FALSE}
packages = installed.packages()
packages = packages[, "Package"]
if (!"kirby21.t1" %in% packages) {
  devtools::install_github("muschellij2/kirby21.t1")
}
```

## Loading Data

We will use the `get_image_filenames_df` function to extract the filenames on our hard disk for the T1 image.  

```{r data}
library(kirby21.t1)
t1_fname = get_t1_filenames(
  ids = 113, 
  visits = 1)
if (is.null(t1_fname)) {
  dl = TRUE
} else {
  dl = !file.exists(t1_fname)
}
if (dl) {
  url = paste0("https://raw.githubusercontent.com/muschellij2/kirby21.t1", "/master/inst/visit_1/113/113-01-T1.nii.gz")
  t1_fname = tempfile(fileext = ".nii.gz")
  download.file(url, t1_fname)
}
```

# T1 image

Let's take a look at the T1-weighted image.  

```{r t1_plot}
t1 = readnii(t1_fname)
ortho2(t1)
```

Here we see the brain and other parts of the image are present.  Most notably, the neck of the subject was imaged.  Sometimes this can cause problems with segmentation and image registration.  

# Attempt 1: Brain Extraction of T1 image using BET

Here we will use FSL's Brain Extraction Tool (BET) to extract the brain tissue from the rest of the image.  

```{r t1_naive_ss, cache = FALSE}
if (have.fsl()) {
  ss_naive = fslbet(infile = t1_fname)
}
```

```{r t1_naive_plot, cache = FALSE}
if (have.fsl()) {
  ortho2(ss_naive)
}

```

We see that naively, BET does not perform well for this image.

# Brain Extraction of T1 image using BET

Here we will use FSL's Brain Extraction Tool (BET) to extract the brain tissue from the rest of the image.  We use the modification of BET in `extrantsr`, which is called through `fslbet_robust`.  In `fslbet_robust`, the image is corrected using the N4 inhomogeneity correction.  The neck of the T1 image is then removed and then BET is run, the center of gravity (COG) is estimated, and BET is run with this new COG.  We used a procedure where the neck is removed in 2 registration steps, which is more robust than just the one (which is the default).

We will do the bias correction explicitly here:

```{r t1_ss, cache = FALSE}
n4img = bias_correct(
  t1_fname, 
  correction = "N4", 
  retimg = FALSE,
  outfile = tempfile(fileext = ".nii.gz"),
  verbose = FALSE)
if (have.fsl()) {
  template.file = fslr::mni_fname(brain = TRUE)
  template.mask = fslr::mni_fname(brain = TRUE, mask = TRUE)
  removed_neck = extrantsr::double_remove_neck(
    n4img,
    template.file = template.file,
    template.mask = template.mask)
  ortho2(removed_neck)
  double_ortho(t1, removed_neck)
}
```

Let's look at the skull-stripped image after doing all these steps.

```{r, echo = TRUE, eval = FALSE}
if (have.fsl()) {
  ss = extrantsr::fslbet_robust(t1_fname, 
    remover = "double_remove_neck", recog = FALSE)
  ortho2(ss)
}
```

```{r, echo = FALSE, eval = TRUE}
if (have.fsl()) {
  ss = fslbet(removed_neck, retimg = TRUE, verbose = FALSE)
  ortho2(ss)
}
```

Here we see the skull-stripped image.  But did we drop out "brain areas"?

```{r t1_ss_plot2, cache = FALSE}
if (have.fsl()) {
  alpha = function(col, alpha = 1) {
    cols = t(col2rgb(col, alpha = FALSE)/255)
    rgb(cols, alpha = alpha)
  }      
  ortho2(t1, ss > 0, col.y = alpha("red", 0.5))
}
```

We can again use `dropEmptyImageDimensions` to remove extraneous slices, which helps with reducing storage of the image on disk, zooming in while plotting, and may aid registration.  

```{r t1_ss_red, cache = FALSE}
if (have.fsl()) {
  ss_red = dropEmptyImageDimensions(ss)
  ortho2(ss_red)
}
```

Again, we can see the zoomed-in view of the image now.